\large Taking uncertainty in word embedding bias estimation seriously - a Bayesian approach

### 2021-08-02 presentationESSLLI21/presentationESSLLI.tex
Working on the presentation

  pdftitle={Taking uncertainty in word embedding bias estimation seriously ~a Bayesian approach},

@@

\title{\large Taking uncertainty in word embedding bias estimation
seriously ~a Bayesian approach}
\author{Alicja Dobrzeniecka \& Rafal Urbaniak
\footnotesize \newline (LoPSE research group, University of Gdansk)}
\date{ExpSem2021, ESSLLI}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias}{}
\begin{block}{Word embeddings}
\protect\hypertarget{word-embeddings}{}
\begin{itemize}
\item
  Representation of words (or words in contexts) with vectors of real
  numbers
\item
  Built to predict the probability of co-occurence
\end{itemize}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-3}{}
\begin{block}{Stereotypical lists}
\protect\hypertarget{stereotypical-lists}{}
\footnotesize

@@

\normalsize
\end{block}

\begin{block}{Visual example}
\protect\hypertarget{visual-example}{}
\vspace{1mm}
\footnotesize

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-4}{}
\begin{block}{Example: direct bias}
\protect\hypertarget{example-direct-bias}{}
\begin{itemize}
\tightlist
\item
  The gender bias of a word \(w\) is its projection on the gender
  direction
  \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
\end{itemize}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-5}{}
\begin{block}{Example: Word Embedding Association Test (WEAT)}
\protect\hypertarget{example-word-embedding-association-test-weat}{}
\begin{align*}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\\
WEAT(A,B) & = \frac{
\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
}{
\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
}
\end{align*}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-6}{}
\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac}{}
\begin{align*}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align*}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-7}{}
\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
\footnotesize 
\begin{table}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-8}{}
\begin{block}{Known challenges}
\protect\hypertarget{known-challenges}{}
\begin{itemize}
\item
  Gender-direction might be an indicator of bias, but is insufficient.
  After debiasing other non-gendered words can remain in biased
  relations (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
  Goldberg, 2019})
\item
  Methods which involve analogies and their evaluations by human users
  on Mechanical Turk are unreliable
  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Cosine distance and word connection visualization}
\protect\hypertarget{cosine-distance-and-word-connection-visualization}{}
\begin{block}{Analysis of word ``muslim''}
\protect\hypertarget{analysis-of-word-muslim}{}
\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Cosine distance and word connection visualization}
\protect\hypertarget{cosine-distance-and-word-connection-visualization-1}{}
\begin{block}{Analysis of word ``priest''}
\protect\hypertarget{analysis-of-word-priest}{}
\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Bayesian model}
\protect\hypertarget{bayesian-model}{}
\end{frame}

\begin{frame}{HPDIs}
\protect\hypertarget{hpdis}{}
\begin{block}{Further work}
\protect\hypertarget{further-work}{}
\begin{itemize}
\tightlist
\item
  downstream tasks
\end{itemize}
\end{block}
\end{frame}

>>>>

  pdftitle={Taking uncertainty in word embedding bias estimation seriously - a Bayesian approach},

@@

\title{\large Taking uncertainty in word embedding bias estimation
seriously - a Bayesian approach}
\author{Alicja Dobrzeniecka \& Rafal Urbaniak
\footnotesize \newline (LoPSE research group, University of Gdansk)}
\date{ExpSem2021, ESSLLI}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias}{}
\begin{block}{Word embeddings}
\protect\hypertarget{word-embeddings}{}
\begin{itemize}
\item
  Representation of words with vectors of real numbers
\item
  Built to predict the probability of co-occurence
\end{itemize}

@@

\begin{frame}{Cosine-based measures of bias - visual example}
\protect\hypertarget{cosine-based-measures-of-bias---visual-example}{}
\begin{block}{Stereotypical lists}
\protect\hypertarget{stereotypical-lists}{}
\footnotesize

@@

\normalsize

\vspace{1mm}
\footnotesize

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-3}{}
\begin{block}{Example: direct bias}
\protect\hypertarget{example-direct-bias}{}
\begin{itemize}
\tightlist
\item
  The gender bias of a word \(w\) is its projection on the gender
  direction
  \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
\end{itemize}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-4}{}
\begin{block}{Example: Word Embedding Association Test (WEAT)}
\protect\hypertarget{example-word-embedding-association-test-weat}{}
\begin{align*}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\\
WEAT(A,B) & = \frac{
\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
}{
\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
}
\end{align*}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-5}{}
\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac}{}
\begin{align*}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align*}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-6}{}
\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
\footnotesize 
\begin{table}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-7}{}
\begin{block}{Known challenges}
\protect\hypertarget{known-challenges}{}
\begin{itemize}
\item
  Gender-direction might be an indicator of bias, but is insufficient.
  After debiasing other non-gendered words can remain in biased
  relations (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
  Goldberg, 2019})
\item
  Methods which involve analogies and their evaluations by human users
  on Mechanical Turk are unreliable
  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Cosine distance and word connection visualization}
\protect\hypertarget{cosine-distance-and-word-connection-visualization}{}
\begin{block}{Analysis of a word ``muslim''}
\protect\hypertarget{analysis-of-a-word-muslim}{}
\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Cosine distance and word connection visualization}
\protect\hypertarget{cosine-distance-and-word-connection-visualization-1}{}
\begin{block}{Analysis of a word ``priest''}
\protect\hypertarget{analysis-of-a-word-priest}{}
\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Advantages of including uncertainty}
\protect\hypertarget{advantages-of-including-uncertainty}{}
\begin{itemize}
\tightlist
\item
  It enables one to directly observe the influence of sample sizes
\item
  It may influence risk assessment and decision making
\item
  \ldots{}
\end{itemize}
\end{frame}

\begin{frame}{Bayesian model}
\protect\hypertarget{bayesian-model}{}
\begin{block}{Choosing priors}
\protect\hypertarget{choosing-priors}{}
\begin{center}\includegraphics[width=1\linewidth]{presentationESSLLI_files/figure-beamer/priorsVis-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Dataset-level HPDIs coefficients}
\protect\hypertarget{dataset-level-hpdis-coefficients}{}
\begin{block}{Religion coefficients}
\protect\hypertarget{religion-coefficients}{}
\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.55\textwidth}
  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
   \end{minipage}
  \end{figure}

\end{center}
\end{block}
\end{frame}

\begin{frame}{Dataset-level HPDIs coefficients}
\protect\hypertarget{dataset-level-hpdis-coefficients-1}{}
\begin{block}{Gender coefficients}
\protect\hypertarget{gender-coefficients}{}
\begin{center}
\begin{figure}[!htb]\centering
  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
\end{minipage}

\end{figure}

\end{center}
\end{block}
\end{frame}

\begin{frame}{Uncertainty included in bias detection}
\protect\hypertarget{uncertainty-included-in-bias-detection}{}
\begin{block}{Closer look at Gender class}
\protect\hypertarget{closer-look-at-gender-class}{}
\includegraphics[width=8cm]{../images/visGenderReddit.png}
\end{block}
\end{frame}

\begin{frame}{Summary of the research}
\protect\hypertarget{summary-of-the-research}{}
\begin{itemize}
\tightlist
\item
\end{itemize}
\end{frame}

\begin{frame}{Further work}
\protect\hypertarget{further-work}{}
\begin{itemize}
\tightlist
\item
  including contrasts in Bayesian model
\item
  downstream tasks
\item
  applying uncertainty to WEAT metric
\item
  testing AIT dataset
\end{itemize}
\end{frame}
